---
title: "Peer Assessment 1"
author: "[Miguel M Fonseca](mailto:mig.m.fonseca@gmail.com)"
date: "17 Apr 2015"
output: 
    html_document:
    toc: true
    toc_depth : 5
---
In this file you will find my report for the *Peer Assessment 1* of the [Reproducible Research][a] course offered by [Coursera][b]

[a]: https://www.coursera.org/course/repdata "Reproducible Research"
[b]: https://www.coursera.org/ "Coursera"


### Loading and preprocessing the data
In this section, I wrote the r code needed to load and preprocess the data to be used in this assignment.

```{r, echo = TRUE}
setwd("/Users/mmfonseca/Documents/work/cursos/2015_Reproducible_Research/Peer_Assessment_1")

## loading the data
ActivityData <- read.csv("activity.csv")
```

The table is already in a suitable format for the analysis.
Check, for example, the first 10 lines of the table (blank cells correspond to "NA"):
```{r, echo = TRUE, results="asis"}
library(xtable)
xt <- xtable(head(ActivityData))
print(xt, type = "html")
```

### What is mean total number of steps taken per day?

*1. Calculate the total number of steps taken per day*
```{r, echo = TRUE, results="asis"}
## In the following command, the number of steps taken per day will be summed up
StepsPerDay <- aggregate(steps ~ date, data = ActivityData, sum)

## just to show how the table looks like
xt <- xtable(head(StepsPerDay))
print(xt, type = "html")
```
<br>

*2. Make a histogram of the total number of steps taken each day*
```{r, echo = TRUE, results="asis"}
## Draw the histogram
hist(StepsPerDay$steps, xlab="total number of steps", 
     main="Histogram of total number of steps taken each day", 
     col="lightblue")
```
<br>

*3. Calculate and report the mean and median of the total number of steps taken per day*
```{r, echo = TRUE, results="asis"}
## For this step, we can use the dataframe created previously (StepsPerDay)
## Note that missing data will be ignored
Mean_StepsPerDay <- mean(StepsPerDay$steps)

## Printing the mean value
show(Mean_StepsPerDay)

## Likewise, we can use the function "media()" to calculate the median of the total number
## of steps taken per day
## Note that missing data will be ignored
Median_StepsPerDay <- median(StepsPerDay$steps)

## Printing the median value
show(Median_StepsPerDay)
```

> The mean of the total number of steps taken per day is **10766.19** and the median of the dataset is **10765**.
<br><br>

### What is the average daily activity pattern?
*1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)*
```{r, echo = TRUE, results="asis"}
AverageStepsPerInterval <- aggregate(steps ~ interval, data = ActivityData, mean)
plot(AverageStepsPerInterval$steps, type="l", lwd=2, 
     ylab="Average Number of Steps", 
     xlab="Daily 5-minute interval",
     main="Time Series Plot"
     )
```
<br>

*2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*
```{r, echo = TRUE, results="asis"}
## To get the max value of a given vector or column, one can use the function which.max
## this function will not report the max value per se, 
## instead it will report the index of the 5-minute interval that, on average, contains the maximum number of steps:
which.max(AverageStepsPerInterval$steps)
```
<br>

So, the above mentioned index interval (or row number), must corresponde to a specific interval. To get the name of the 5-minute interval, one can use the following:
```{r, echo = TRUE, results="asis"}
AverageStepsPerInterval$interval[which.max(AverageStepsPerInterval$steps)]
```

> The 5-minute interval that contains the maximum number of steps (on average across all the days in the dataset is the one named **835**.

### Imputing missing values
*1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)*
```{r, echo = TRUE, results="asis"}
nrow(ActivityData[ActivityData$steps == "NA",])
```

*2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.*

*3. Create a new dataset that is equal to the original dataset but with the missing data filled in.*

I will answer both questions simultaneously using a for loop.
I will use the overall 5-minute interval mean to fill the missing values. To do this I will use the values from the dataframe "AverageStepsPerInterval". Thus, if in the original dataset the value of number of steps is missing, I will substitute it will the average number of steps for that 5-minute interval:

```{r, echo = TRUE, results="asis"}
ActivityData2 <- ActivityData

for (i in 1:nrow(ActivityData)){
    interval <- ActivityData$interval[i]
    nsteps <- ActivityData$steps[i]
    newvalue <- AverageStepsPerInterval[AverageStepsPerInterval$interval == interval,2]
    if (is.na(nsteps)){
        ActivityData2$steps[i] <- newvalue
    }
}
````
<br>

> The new dataframe with the missing values substituted with the specific 5-minute interval average was named `ActivityData2`.
<br>

*4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.*

Drawing the histogram based on the dataset with no missing values...
```{r, echo = TRUE, results="asis"}
StepsPerDay2 <- aggregate(steps ~ date, data = ActivityData2, sum)

hist(StepsPerDay2$steps, xlab="total number of steps", 
     main="Histogram of total number of steps taken each day (no missing values)", 
     col="blue")
```
<br>

And calculating and reporting the mean and median total number of steps taken per day.
Mean:
```{r, echo = TRUE, results="asis"}
Mean_StepsPerDay2 <- mean(StepsPerDay2$steps)
show(Mean_StepsPerDay2)
```
Median:
```{r, echo = TRUE, results="asis"}
Median_StepsPerDay2 <- median(StepsPerDay2$steps)
show(Median_StepsPerDay2)
```
<br>
*Do these values differ from the estimates from the first part of the assignment?*
The mean value does not change, but the median does because I have filled the original dataset with new values that made a change in the sample median.
```{r, echo = TRUE, results="asis"}
## Summarizing the dataset (daily steps average) before missing data was filled in
summary(StepsPerDay$steps)

## and after
summary(StepsPerDay2$steps)
```

*What is the impact of imputing missing data on the estimates of the total daily number of steps?*
There is an impact on the quartiles and the median, but not on the minimum, maximum or mean values.
<br>

### Are there differences in activity patterns between weekdays and weekends?
*1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.*
To create the new factor, I will use the fuction weekdays to the column "date" of the dataset "ActivityData2". The resultant vector will be added to the dataset.
```{r, echo = TRUE, results="asis"}

week_vector <- weekdays(as.Date(ActivityData2$date))

## adding the vector as a new column of ActivityData2
ActivityDataNew <- cbind(ActivityData2,week_vector)
```
